On Optimality of Finite Element Integration

### 2016-02-13 TOMS_Paper/assembly.tex
before removing complete bipartite graphs

\item If $\mathbb{S} = \mathbb{T}$, jump to step (7). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}.
\item There exists a complete bipartite sub-graph $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$, with $\mathbb{S}_j$ and $\mathbb{S}_k$ being partitions of symbols depending on $L_j$ and $L_k$, respectively: take an arbitrary symbol in $\mathbb{S}_j^d \cup \mathbb{S}_k^d$ and apply Strategy~\ref{strategy:ii}.
\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply Strategy~\ref{strategy:ii}. \end{enumerate}
Go back to step (2).

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D. The operation counts are $E(f(z^0, z^1, ...) + IJK \cdot 18)$ (left) and $E(f(z^0, z^1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z^0, z^1, ...)$ representing the operation count for evaluating $z^0, z^1, ...$, and common sub-expressions being counted once. Note that the synthesis in Figure~\ref{code:poisson_b} reaches optimality, apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}

@@

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination. In particular, we define $\theta(m) = \mathrm{\theta}^{pre}(m) - \mathrm{\theta}^{se}(m)$, where $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation to, respectively. Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) < 0$. We return to the issue of deriving $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can now describe the transformation algorithm (Algorithm~\ref{algo:gamma}).

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may not be optimal, since $\theta$ only distinguishes monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials sharing some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a non-optimum. This situation is prevented by construction, because $\Gamma$ exhaustively searches all possible bipartitions for determining an optimum preserving Constraint~\ref{const:TH}\footnote{Note that the problem can be seen as an instance of the well-known Knapsack problem}. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Proposition~\ref{prop:multi-struct} ensures that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (through Strategy~\ref{strategy:ii}). If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration. Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) cannot prune the optimum from the search space.

Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules. Non flop-decreasing factorizations are excluded by construction at step (5). We have to prove that applying in sequence rewrite rules (A), (B), (C), and (D), with applications of rules (C) and (D) that may expose further opportunities for (A) and (B), represents a step forward towards quasi-optimality. 
\begin{description}
\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, we see that setting $V = T$ and $S = S^{nt}$ would simply expose another hit for rule (B). Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ... + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$. From this, we see that applying rule (C) prior to rule (B) may prune the optimum from the search space.
\item[Rule (C)] We interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively. There can be multiple connected components, each of which we denote by $\mathbb{B}^d$. We identify with $lb^d$ the lower bound on the operation count, which we recall is an affine function of the number of products, induced by the symbols in $\mathbb{B}^d$. The factorization in any $\mathbb{B}^d$ is independent of that in other connected components. It therefore suffices to consider an arbitrary $\mathbb{B}^d$. We distinguish the two following possibilities.
\begin{description}
\item[$\mathbb{B^d}$ is complete] We simply have that $lb^d = 1$, which the algorithm reaches through iterative factorization (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$), by jumping back to step (2) at the end of step (6). 
\item[$\mathbb{B^d}$ is not complete] As a consequence of loop linearity and the absence of further hits for rules (A) and (B), $lb^d = {\min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. 
By not choosing the vertex with highest degree (as the rule states),  ...
\end{description}
\end{description}
The algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
\end{enumerate}
\end{proof}

@@

For a fair comparison, small patches (publicly available) were written for compatibility of \texttt{quad}, \texttt{tens}, and \texttt{ufls} with Firedrake. By executing all simulations in Firedrake, we guarantee that matrix insertion and mesh iteration have a fixed cost independent of the transformation system employed. The patches, in particular, adjust the data storage layout (e.g., by generating array of pointers instead of pointer to pointers and by replacing flattened arrays with bi-dimensional ones, as Firedrake expects).

>>>>

\item If $\mathbb{S} = \mathbb{T}$, jump to step (7). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: take $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}.
\item There exists a complete bipartite sub-graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, with $\mathbb{S}_j$ and $\mathbb{S}_k$ being partitions of symbols depending on $L_j$ and $L_k$, respectively: apply Strategy~\ref{strategy:ii} to either the symbols in $\mathbb{S}_j$ or $\mathbb{S}_k$ (choice is arbitrary).
\item $\mathbb{T} = \emptyset$, $S$ is in a cycle (i.e., a path starting and ending in $S$ with at least one edge and no repeated edges or vertices): take $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: the precondition basically discards factorizations that can be ``distributed within larger factorizations''. Further details will be provided in Section~\ref{sec:proof}}.
\end{enumerate}
Go back to step (2).

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D. The operation counts are $E(f(z^0, z^1, ...) + IJK \cdot 18)$ (left) and $E(f(z^0, z^1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z^0, z^1, ...)$ representing the operation count for evaluating $z^0, z^1, ...$, and common sub-expressions being counted once. Note that the synthesis in Figure~\ref{code:poisson_b} reaches optimality apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}

@@

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination. In particular, we define $\theta(m) = \mathrm{\theta}^{pre}(m) - \mathrm{\theta}^{se}(m)$, where $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively. Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) < 0$. We return to the issue of deriving $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can now describe the transformation algorithm (Algorithm~\ref{algo:gamma}).

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may not be optimal, since $\theta$ only distinguishes monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials sharing some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a non-optimum. This situation is prevented by construction, because $\Gamma$ exhaustively searches all possible bipartitions for determining an optimum preserving Constraint~\ref{const:TH}\footnote{Note that the problem can be seen as an instance of the well-known Knapsack problem}. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Proposition~\ref{prop:multi-struct} ensures that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (via Strategy~\ref{strategy:ii}). If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration. Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) cannot prune the optimum from the search space.

Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules. Non flop-decreasing factorizations are excluded by construction at step (5). We first discuss the individual impact of the rewrite rules; then, we explain why applying them in a certain order represents a step forward towards quasi-optimality.

\begin{description}
\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, we see that setting $V = T$ and $S = S^{nt}$ would simply expose another hit for rule (B). Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ... + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$. From this, we evince that applying rule (D) prior to rule (B) may prune the optimum from the search space.
\item[Rule (C)] The fact that $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$ is complete implies that a sub-expression can be implemented by only one multiplication at every inner loop iteration. For example, the expression $S^0_j S^0_k + S^0_j S^1_k + S^1_j S^0_k + S^1_j S^1_k$ can first be recast (factorization) as $(S^0_j + S^1_j) (S^0_k + S^1_k)$ and finally (code motion) as $P_j P_k$. Rule (C) will apply Strategy~\ref{strategy:ii} ${\min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$ times, thus reaching the objective. 
\item[Rule (D)] The precondition implies that whatever factorization is chosen, no symbols will be eradicated from the expression. The rule then acts conservatively: the absence of a cycle through $S$ simply implies that $S$ will be hoisted through other applications of Strategy~\ref{strategy:ii}, so there is no need to perform a factorization.
\end{description}

Instances of (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid regardless of how the applications of Strategy~\ref{strategy:ii} are scheduled. (C), similarly, guarantees a single multiplication for a whole set of symbols. From this, we deduce that what Algorithm~\ref{algo:sharing-elimination} actually has to guarantee is a partial ordering: (A), (B), and (C) can be applied in arbitrary order, whereas (D), as explained above, must be considered only when none of the other rules can be hit. Rule (D) is clearly the most expensive from a computational point of view, because it must prune all choices that will not lead to the optimum. However, this rule becomes relevant only in extremely complex forms, in which extensive use of differential operators is made.

Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
\end{enumerate}
\end{proof}

@@

For a fair comparison, small patches (publicly available) were written for the compatibility of \texttt{quad}, \texttt{tens}, and \texttt{ufls} with Firedrake. By executing all simulations in Firedrake, we guarantee that both matrix insertion and mesh iteration have a fixed cost, independent of the transformation system employed. The patches adjust the data storage layout to what Firedrake expects (e.g., by generating an array of pointers instead of a pointer to pointers, by replacing flattened arrays with bi-dimensional ones).
